```{r}
data <- read.csv("C:/Users/obero/Downloads/projectdata3.csv")

# To view the data
num_rows <- nrow(data)
num_columns <- ncol(data)

# Print the details
cat("Number of rows in the dataset:", num_rows, "\n")
cat("Number of columns in the dataset:", num_columns)
```

```{r}
library(readr)

# Read the original CSV file using base R 
data <- read.csv("C:/Users/obero/Downloads/projectdata3.csv")
                                 
# Select only the specified columns
selected_data <- data[, c("V2026", "V2033", "V2034", "V2036", "V2040", "V2072", "V2073")]

# Remove rows with NA values in any of the selected columns
selected_data_clean <- na.omit(selected_data)

# Rename the columns in the cleaned data
colnames(selected_data_clean) <- c("HouseholdIncome", "PrincipalPersonAge", "MaritalStatus", 
                                   "PrincipalPersonSex", "PrincipalPersonRace", 
                                   "NumHHMembersUnder12", 
                                   "NumCrimeIncidentReports")

# Write the cleaned data to a new CSV file
write.csv(selected_data_clean, "C:/Users/obero/Downloads/Regression_data.csv", row.names = FALSE)

```


```{r}
selected_data_clean <- read.csv("C:/Users/obero/Downloads/Regression_data.csv")
```

```{r}
head(selected_data_clean)
```

```{r paged.print=TRUE}
# Exploratory Data Analysis + Additional Data Cleaning

# Re-categorize HouseholdIncome into 5 categories
selected_data_clean$HouseholdIncome <- with(selected_data_clean, ifelse(HouseholdIncome=="(01) Less than $5,000", "< 5000",
                                    ifelse(HouseholdIncome=="(02) $5,000-$7,499" | HouseholdIncome=="(03) $7,500-$9,999" | HouseholdIncome=="(04) $10,000-$12,499" | HouseholdIncome=="(05) $12,500-$14,999" | HouseholdIncome=="(06) $15,000-$17,499" | HouseholdIncome=="(07) $17,500-$19,999" | HouseholdIncome=="(08) $20,000-$24,999", "5000-25000",
                                           ifelse(HouseholdIncome=="(09) $25,000-$29,999" | HouseholdIncome=="(10) $30,000-$34,999" | HouseholdIncome=="(11) $35,000-$39,999" | HouseholdIncome=="(12) $40,000-$49,999", "25000-50000",
                                                     ifelse(HouseholdIncome=="(13) $50,000-$74,999", "50000-75000",
                                                            ifelse(HouseholdIncome=="(14) $75,000 and over", "75000+", NA))))))
#selected_data_clean <- na.omit(selected_data_clean)

# Summary of the cleaned and encoded data

selected_data_clean$HouseholdIncome <- factor(selected_data_clean$HouseholdIncome,
                                              levels = c("< 5000", "5000-25000", "25000-50000", "50000-75000", "75000+"),
                                              labels = c("Less than $5,000", "$5,000-$25,000", "$25,000-$50,000", "$50,000-$75,000", "Over $75,000"))

# Factor Encoding for Marital Status, Principal Person Sex, and Principal Person Race
selected_data_clean$MaritalStatus <- factor(selected_data_clean$MaritalStatus,
                                            levels = c("(1) Married", "(2) Widowed", "(3) Divorced", "(4) Separated", "(5) Never married"),
                                            labels = c("1", "2", "3", "4", "5"))


selected_data_clean$PrincipalPersonSex <- factor(selected_data_clean$PrincipalPersonSex,
                                                 levels = c("(1) Male", "(2) Female"),
                                                 labels = c("1", "2"))

selected_data_clean$PrincipalPersonRace <- factor(selected_data_clean$PrincipalPersonRace,
                                                  levels = c("(1) White", "(2) Black", "(3) American Indian, Aleut, Eskimo", "(4) Asian, Pacific Islander"),
                                                  labels = c("1", "2", "3", "4"))
summary(selected_data_clean)
```


```{r paged.print=TRUE}
table(selected_data_clean$MaritalStatus)

```


```{r}
#Exploratory Data Analysis
# Load necessary library for plotting
library(ggplot2)

ggplot(selected_data_clean, aes(x = MaritalStatus)) +
  geom_bar(fill = "steelblue") +
  scale_x_discrete(labels = c("1" = "Married", 
                              "2" = "Widowed", 
                              "3" = "Divorced", 
                              "4" = "Separated", 
                              "5" = "Never Married")) +
  labs(title = "Distribution of Marital Status",
       subtitle = "Analysis of Marital Status Frequency",
       x = "Marital Status",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Assuming 'PrincipalPersonAge' is a column in your dataframe `selected_data_clean`
# Define the breaks for the age ranges you want to use
age_breaks <- seq(from = floor(min(selected_data_clean$PrincipalPersonAge, na.rm = TRUE)/10)*10, 
                  to = ceiling(max(selected_data_clean$PrincipalPersonAge, na.rm = TRUE)/10)*10, by = 10)

# Create the histogram with defined breaks using updated ggplot2 syntax
ggplot(selected_data_clean, aes(x = PrincipalPersonAge)) +
  geom_histogram(breaks = age_breaks, 
                 fill = "blue", 
                 color = "black", 
                 aes(y = after_stat(count))) +
  scale_x_continuous(breaks = age_breaks) +
  labs(title = "Distribution of Age", 
       x = "Age Range", 
       y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))



ggplot(selected_data_clean, aes(x = PrincipalPersonSex)) +
  geom_bar(fill = "coral") +
  scale_x_discrete(labels = c("1" = "Male", "2" = "Female")) +
  labs(title = "Distribution of Principal Person Sex",
       subtitle = "Comparison of Male and Female Counts",
       x = "Sex",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 0))  # Adjust angle if needed

ggplot(selected_data_clean, aes(x = PrincipalPersonRace)) +
  geom_bar(fill = "lightgreen") +
  scale_x_discrete(labels = c("1" = "White", 
                              "2" = "Black", 
                              "3" = "American Indian, Aleut, Eskimo", 
                              "4" = "Asian, Pacific Islander")) +
  labs(title = "Distribution of Principal Person Race",
       subtitle = "Diversity in Racial Demographics",
       x = "Race",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


ggplot(selected_data_clean, aes(x = HouseholdIncome, fill = MaritalStatus)) +
  geom_bar(position = "fill") +
  scale_fill_discrete(name = "Marital Status",
                      labels = c("1" = "Married",
                                 "2" = "Widowed",
                                 "3" = "Divorced",
                                 "4" = "Separated",
                                 "5" = "Never Married")) +
  labs(title = "Income Group by Marital Status",
       x = "Income Group",
       y = "Proportion") +
  theme_minimal()


ggplot(selected_data_clean, aes(x = PrincipalPersonRace, fill = PrincipalPersonSex)) +
  geom_bar(position = "dodge") +
  scale_x_discrete(labels = c("1" = "White", 
                              "2" = "Black", 
                              "3" = "American Indian, Aleut, Eskimo", 
                              "4" = "Asian, Pacific Islander")) +
  scale_fill_discrete(name = "Sex", 
                      labels = c("1" = "Male", "2" = "Female")) +
  labs(title = "Distribution of Principal Person Sex by Race", 
       x = "Race", 
       y = "Count", 
       fill = "Sex") +
  theme_minimal()

selected_data_clean$PrincipalPersonRace <- factor(selected_data_clean$PrincipalPersonRace,
                                                  levels = c(1, 2, 3, 4),
                                                  labels = c("White", "Black", "American Indian/Aleut/Eskimo", "Asian/Pacific Islander"))

# Now the ggplot code
ggplot(selected_data_clean, aes(x = HouseholdIncome, fill = PrincipalPersonRace)) +
  geom_bar(position = "dodge", stat = "count") +
  scale_fill_manual(name = "Race",
                    values = c("White" = "1", "Black" = "2", "American Indian/Aleut/Eskimo" = "3", "Asian/Pacific Islander" = "4"),
                    labels = c("White", "Black", "American Indian/Aleut/Eskimo", "Asian/Pacific Islander")) +
  labs(title = "Household Income by Race",
       x = "Income Group",
       y = "Count") +
  theme_minimal()

# Replace "color1", "color2", etc., with your chosen colors

```

```{r}

library(ggplot2)
library(ggmosaic)
library(corrplot)

# Assuming you have some numerical variables for correlation analysis
numerical_data <- selected_data_clean[, sapply(selected_data_clean, is.numeric)]
cor_matrix <- cor(numerical_data, use = "complete.obs")
corrplot(cor_matrix, method = "circle")

```
```{r}

unique(selected_data_clean$PrincipalPersonRace)

```


```{r}
# Hypothesis Testing

# Household Income 

# H0: Household income has no significant effect on the number of crime incident reports.
# H1: Household income significantly affects the number of crime incident reports.

model_income <- lm(NumCrimeIncidentReports ~ HouseholdIncome, data = selected_data_clean)
summary(model_income)
anova(model_income)


# As the p-value is less than 0.05 we reject the null hypothesis. Hence, household income significantly affects the number of crime incident reports.


# Principal Person Age

# H0: The age of the principal person has no significant effect on the number of crime incident reports.
# H1: The age of the principal person significantly affects the number of crime incident reports.

model_age <- lm(NumCrimeIncidentReports ~ PrincipalPersonAge, data = selected_data_clean)
summary(model_age)
anova(model_age)

# As the p-value is less than 0.05 we reject the null hypothesis. Hence, age significantly affects the number of crime incident reports.


# Principal Person Marital Status

# H0: Marital status has no significant effect on the number of crime incident reports.
# H1: Marital status significantly affects the number of crime incident reports.

model_marital_status <- lm(NumCrimeIncidentReports ~ MaritalStatus, data = selected_data_clean)
summary(model_marital_status)
anova(model_marital_status)

# As the p-value is less than 0.05 we reject the null hypothesis. Hence, marital status significantly affects the number of crime incident reports.

# Principal Person Sex

# H0: The sex of the principal person has no significant effect on the number of crime incident reports.
# H1: The sex of the principal person significantly affects the number of crime incident reports.

model_sex <- lm(NumCrimeIncidentReports ~ PrincipalPersonSex, data = selected_data_clean)
summary(model_sex)
anova(model_sex)

# As the p-value is less than 0.05 we reject the null hypothesis. Hence, sex of the principal person significantly affects the number of crime incident reports.

# Principal Person Race

# H0: The race of the principal person has no significant effect on the number of crime incident reports.
# H1: The race of the principal person significantly affects the number of crime incident reports.

model_race <- lm(NumCrimeIncidentReports ~ PrincipalPersonRace, data = selected_data_clean)
summary(model_race)
anova(model_race)

# As the p-value is less than 0.05 we reject the null hypothesis. Hence, race of the principal person significantly affects the number of crime incident reports.


# Number of Household Members Younger Than 12 Years

# H0: The number of household members younger than 12 years has no significant effect on the number of crime incident reports.
# H1: The number of household members younger than 12 years significantly affects the number of crime incident reports.

model_household_members_under_12 <- lm(NumCrimeIncidentReports ~ NumHHMembersUnder12, data = selected_data_clean)
summary(model_household_members_under_12)
anova(model_household_members_under_12)

# As the p-value is less than 0.05 we reject the null hypothesis. Hence, the number of household members less than 12 years significantly affects the number of crime incident reports.
```


```{r}

#Model Making

# Load necessary library
library(ggplot2)


#selected_data_clean$HouseholdIncome <-factor(selected_data_clean$NumCrimeIncidentReports)
  # Assuming 'data' is your dataframe and it has been cleaned and prepared
model <- lm(NumCrimeIncidentReports ~ ., data = selected_data_clean)

# Summary of the model
summary(model)


# lets 
```
The initial model fetched a very low value of R - squared adjusted
Lets try different algorithms like poisson regression.
```{r}
# Poisson Regression
poisson_model <- glm(NumCrimeIncidentReports ~ ., data = selected_data_clean, family = poisson())
summary(poisson_model)
```
we transformed the data by first removing the intercept which fetched us a value of 0.6 r-squared adjusted 

```{r}
# without log transformation
mlm_model <- lm(NumCrimeIncidentReports ~ .-1, data = selected_data_clean)
summary(mlm_model)
```
After this, we checked with the log transformation of the response variable which fetched a value of 0.90
```{r}

# with log transformation
selected_data_clean$Log_NumCrimeIncidentReports <- log(selected_data_clean$NumCrimeIncidentReports + 1)

selected_data_clean$MaritalStatus <- as.factor(selected_data_clean$MaritalStatus)
selected_data_clean$PrincipalPersonSex <- as.factor(selected_data_clean$PrincipalPersonSex)
selected_data_clean$PrincipalPersonRace <- as.factor(selected_data_clean$PrincipalPersonRace)

# Fit the model with all predictors and check for interactions
combined_model <- lm(Log_NumCrimeIncidentReports ~., data = selected_data_clean)
summary(combined_model)


```

We perform the model diagnostics and check whether the model is highly overfitted
Looking at the plot of residual vs fitted values, we observe non-normality, 

We identify that regression may not be the right approach to handle our problem statement.


```{r}
library(MASS)  # for boxcox function

# Ensure positive values for the response variable
selected_data_clean <- selected_data_clean[selected_data_clean$NumCrimeIncidentReports > 0, ]

# Apply Box-Cox transformation
bc <- boxcox(NumCrimeIncidentReports ~ ., data = selected_data_clean)
lambda <- bc$x[which.max(bc$y)]
selected_data_clean$BC_NumCrimeIncidentReports <- (selected_data_clean$NumCrimeIncidentReports^lambda - 1) / lambda

# Fit the model with the Box-Cox transformed response
model <- lm(BC_NumCrimeIncidentReports ~ ., data = selected_data_clean)
#summary(model)

# Diagnostic plots
par(mfrow=c(2,2))
plot(model)


```
```{r}
# Load necessary libraries
library(car)

# Assuming 'model' is your lm model object
# Calculate Cook's distance
cooksd <- cooks.distance(model)

# Plot Cook's distance
plot(cooksd, pch="*", cex=2, main="Cook's Distance")
abline(h = 4/(nrow(model$model)-length(model$coefficients)), col="red")

# Add the rule of thumb line for Cook's Distance
# Influential points are those for which Cook's distance exceeds this threshold
influential <- as.numeric(names(cooksd)[(cooksd > (4/(nrow(model$model) - length(model$coefficients))))])

# Check for leverage points
leverage_points <- which(hatvalues(model) > (2 * mean(hatvalues(model))))

# Bonferroni p-value for outlier detection
outlier_test <- car::outlierTest(model)

# Combine all results in a list
diagnostics <- list(
  cooks_distance = cooksd,
  influential_points = influential,
  leverage_points = leverage_points,
  outliers = outlier_test
)

# Print the diagnostics
#print(diagnostics)

```



Classification Approach( Out of scope for Course)
```{r}
#Classification Approach
crime_data<- read.csv("C:/Users/obero/Downloads/Regression_data.csv", header = TRUE)

#crime_data

# Ensure the randomForest package is installed
if (!require(randomForest)) install.packages("randomForest")
library(randomForest)

# Convert factors
crime_data$MaritalStatus <- as.factor(crime_data$MaritalStatus)
crime_data$PrincipalPersonSex <- as.factor(crime_data$PrincipalPersonSex)
crime_data$PrincipalPersonRace <- as.factor(crime_data$PrincipalPersonRace)

# Fit Random Forest Model
rf_model <- randomForest(as.factor(NumCrimeIncidentReports) ~ ., data = crime_data)

# Print model summary
print(rf_model)


```
{out of scope of course, we apply the algoirthms of classification(random forest)}



```{r}
# Assuming your Random Forest model is named rf_model
feature_importance <- importance(rf_model)
# Sort the feature importance
sorted_feature_importance <- feature_importance[order(feature_importance[, 1], decreasing = TRUE), ]



```


```{r}
sorted_feature_importance
```

# PrincipalPersonAge (3782.0721): This is the most influential feature in your model. A higher importance score indicates that the age of the principal person has the strongest relationship with the number of crime incident reports. This suggests that different age groups might have varying tendencies or exposure to crime incidents.

#HouseholdIncome (2064.7088): The second most important feature. The income level of the household shows a significant impact on crime incident reports. This could reflect socio-economic factors that influence crime rates or reporting habits.

#NumHHMembersUnder12 (1151.9126): The number of household members under the age of 12 is also a key predictor. This might indicate that households with more children might have different crime reporting patterns or experiences.

#MaritalStatus (1002.6603): This suggests that the marital status of the principal person is an important factor. Different marital statuses might correlate with different lifestyles or residential areas, which in turn might affect crime exposure or reporting.

#PrincipalPersonRace (822.9726): The race of the principal person is also significant. This could point to demographic factors related to crime incidents or reporting, reflecting broader social and environmental contexts.

#PrincipalPersonSex (430.6977): The least influential of the variables listed, but still important. The sex of the principal person has an impact, though less than other variables. This might capture gender-related differences in crime exposure or reporting behaviors.





```{r}
# Assuming sorted_feature_importance is a named vector as shown in your output
sorted_feature_importance <- c(PrincipalPersonAge = 3782.0721, HouseholdIncome = 2064.7088, 
                               NumHHMembersUnder12 = 1151.9126, MaritalStatus = 1002.6603, 
                               PrincipalPersonRace = 822.9726, PrincipalPersonSex = 430.6977)

# Sort the feature importance in decreasing order
sorted_feature_importance <- sort(sorted_feature_importance, decreasing = TRUE)

# Create a bar plot
barplot(sorted_feature_importance,
        main = "Feature Importance in Random Forest Model",
        xlab = "Feature Importance Score",
        ylab = "Features",
        horiz = TRUE, # Horizontal bar plot
        las = 1, # Make labels horizontal
        col = "blue")

```
```{r} 
# Assuming 'data' is your dataframe and it's already been read into R.

# Load the necessary library
library(dplyr)

# Convert all character columns to factors
data <- selected_data_clean %>%
  mutate_if(is.character, as.factor)

# Check the levels of each factor (categorical variable)
sapply(data, function(x) if(is.factor(x)) levels(x))


```

